---
title: "MSA Cluster Modeling"
author: "Robert J. DiNapoli"
date: "8/16/2022"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

# Load Packages
```{r, results='hide', message=FALSE}
library(here)
library(mclust)
library(raster)
library(rgdal)
library(sp)
library(RColorBrewer)
library(PCAtools)
library(sf)
library(tmap)
library(terra)
library(rgeos)
library(spData)
library(knitr)
library(kableExtra)
```

# Import Data
```{r, results='hide', message=FALSE}

#import raw data
dat_raw <- read.csv(here("data/CommunityData-raw-2015-v2.csv"))
#select columns of interest, for this run remove asian_equality and female_equality
dat  <- dat_raw[,c(4:18,20:21)]
#convert to dataframe
dat <- as.data.frame(unclass(dat))
#set rownames to city names
rownames(dat)=dat_raw$ME
# remove any records that have NAs
dat = na.omit(dat)
# convert to z-scores
dat <- as.matrix(scale(dat, center = TRUE, scale = TRUE))
```

# Run PCA on dataset
```{r}
dat_comps <- prcomp(dat, center = F, scale=F) #set center and scale to FALSE because done pre-PCA
#get summary
s <- summary(dat_comps)
kable(s$importance)
#get eigenvalues
#### change this # of columns to match
ev <- data.frame(Component=paste("PC",1:17, sep=""),eigenvalue=dat_comps$sdev^2)
kable(ev)

#write to CSV
write.csv(dat_comps$x,file=here("results/community_loadings.csv"))


```

## Extract components with eigenvalues > 1
```{r}
dat_pcs <- dat_comps$x[,1:5]
```

## Plot loadings
```{r}
#plot loadings of pc1 and 2
plot(dat_comps$x[,1],dat_comps$x[,2], 
     xlab=paste("PCA 1 (", round(s$importance[2,1]*100, 1), "%)", sep = ""), 
     ylab=paste("PCA 2 (", round(s$importance[2,2]*100, 1), "%)", sep = ""), 
     pch=16, col="blue", cex=0.5)
abline(v=0, lwd=2, lty=2)
abline(h=0, lwd=2, lty=2)
#get loadings
l.x <- dat_comps$rotation[,1]*10
l.y <- dat_comps$rotation[,2]*10
arrows(x0=mean(dat_comps$x[,1]), x1=l.x, y0=mean(dat_comps$x[,2]), y1=l.y, col="black", length=0.15, lwd=1.5)
# Label position
l.pos <- l.y # Create a vector of y axis coordinates
lo <- which(l.y < 0) # Get the variables on the bottom half of the plot
hi <- which(l.y > 0) # Get variables on the top half
# Replace values in the vector
l.pos <- replace(l.pos, lo, "1")
l.pos <- replace(l.pos, hi, "3")
text(l.x, l.y, labels=row.names(dat_comps$rotation), col="black", pos=l.pos, cex=1.25)

#plot loadings of pc 3 and 4
plot(dat_comps$x[,3],dat_comps$x[,4], 
     xlab=paste("PCA 3 (", round(s$importance[2,3]*100, 1), "%)", sep = ""), 
     ylab=paste("PCA 4 (", round(s$importance[2,4]*100, 1), "%)", sep = ""), 
     pch=16, col="blue", cex=0.5)
abline(v=0, lwd=2, lty=2)
abline(h=0, lwd=2, lty=2)

#get loadings
l.x <- dat_comps$rotation[,3]*10
l.y <- dat_comps$rotation[,4]*10
arrows(x0=mean(dat_comps$x[,3]), x1=l.x, y0=mean(dat_comps$x[,4]), y1=l.y, col="black", length=0.15, lwd=1.5)
# Label position
l.pos <- l.y # Create a vector of y axis coordinates
lo <- which(l.y < 0) # Get the variables on the bottom half of the plot
hi <- which(l.y > 0) # Get variables on the top half
# Replace values in the vector
l.pos <- replace(l.pos, lo, "1")
l.pos <- replace(l.pos, hi, "3")
text(l.x, l.y, labels=row.names(dat_comps$rotation), col="black", pos=l.pos, cex=1.25)
```
## Summary of PC loadings
```{r}
library(broom)
print(dat_comps)
summary(dat_comps)
# information about PCs
tidy(dat_comps, "pcs")
```

# Run Cluster GMM

Assumptions: - MSAs form clusters characterized by a multivariate distribution
             - Model forms: shape, volume, orientation
             - GMM fits a series of models with different forms and numbers of clusters
             - Models with highest probability and fewest parameters selected as most optimal
               - Based on Bayesian Information Criterion (BIC)

```{r}
#run GMM on 1-20 clusters
dat_pcs_mc <- Mclust(dat_pcs, G=c(1:20))
#print summary of model fit
summary(dat_pcs_mc)
```

Cluster assignments
```{r}
plot(dat_pcs_mc, what = c("classification"))
```
Uncertainty in assignments
```{r}
write.csv(dat_pcs_mc$uncertainty,file=here("results/community_group_uncertainty.csv"))
plot(dat_pcs_mc, what = "uncertainty")
uncerPlot(z = dat_pcs_mc$z)
```

# Model comparison
```{r}
#plot BIC scores of different models
plot(dat_pcs_mc, what='BIC')
#zoom in on best-fitting models
plot(dat_pcs_mc, what='BIC', ylim=c(-15200,-14600))

#model type VEI and VEE have similar BIC scores for 8-9 clusters
#do additional model comparisons
m_VEI_8 <- Mclust(dat_pcs, G=8, modelNames="VEI")
m_VEE_8 <- Mclust(dat_pcs, G=8, modelNames="VEE")
m_VEI_9 <- Mclust(dat_pcs, G=9, modelNames="VEI")
m_VEI_10 <- Mclust(dat_pcs, G=10, modelNames="VEI")
m_VEI_11 <- Mclust(dat_pcs, G=11, modelNames="VEI")
m_VEE_9 <- Mclust(dat_pcs, G=9, modelNames="VEE")
m_VEE_10 <- Mclust(dat_pcs, G=10, modelNames="VEE")
m_VEE_11 <- Mclust(dat_pcs, G=11, modelNames="VEE")
#extract BIC scores
BICs<-c(m_VEI_8$BIC[1],m_VEE_8$BIC[1],m_VEI_9$BIC[1],m_VEE_9$BIC[1],m_VEI_10$BIC[1],m_VEE_10$BIC[1],m_VEI_11$BIC[1],m_VEE_11$BIC[1])
#calculate change in BIC score, since in this method the goal to to maximize BIC
#we calculate change from highest scoring model
delta_BIC <- max(BICs) - BICs
#calculate BIC weights
w_BIC <- round(exp(-0.5*delta_BIC)/sum(exp(-0.5*delta_BIC)), digits=3)
#make table of results
results_table <- cbind.data.frame(clusters=c("VEI_8","VEE_8","VEI_9","VEE_9","VEI_10","VEE_10","VEI_11","VEE_11"),BIC=BICs,delta_BIC,weight=w_BIC)
#order by delta
results_table <- results_table[order(delta_BIC),] 
rownames(results_table)<-NULL
#print table
kable(results_table[1:4], caption = "Model Comparison")

## providing some output about the model. It seems that VEI 11 is the best model. 
## MODEL CHOSEN FOR REST OF ANALYSES

output <- clustCombi(data = dat_pcs, modelName = "VEI", G = 11)
# plot the hierarchy of combined solutions
plot(output, what = "classification")
# plot some "entropy plots" which may help one to select the number of classes
plot(output, what = "entropy")
# plot the tree structure obtained from combining mixture components
plot(output, what = "tree")
head( output$combiz[[output$MclustOutput$G]] )

## output the probabilities by cluster.
write.csv(output$combiz[[output$MclustOutput$G]],file=here("results/cluster_probabilities.csv"))
```

# Extract and map cluster classifications
```{r}
#extract classifications, e.g., which city belongs to which cluster, export to csv and reload

### Change this value to pick a different model
## currently the best solution is VEI with 11 clusters.

write.csv(m_VEI_11$classification,file=here("results/cluster_assignment_pca.csv"))
data<- read.csv(here('results/cluster_assignment_pca.csv'))

#import shapefile
msa_Boundary <-readOGR(here("data/MSA"),"tl_2015_us_cbsa") 
#merge them
merged <- merge(msa_Boundary,data,by.x="NAME",by.y="X")
#remove any cases MSAs with no cluster assignments
merged_clean <- merged[!is.na(merged$x),]
#convert x to factor
merged_clean$Cluster <- as.factor(merged_clean$x)
#convert to sf object
cluster_sf <- st_as_sf(merged_clean)
#make us map
us_map <- tm_shape(us_states) + tm_borders()
#combine them
cluster_map <- tm_shape(cluster_sf) + tm_fill(col="Cluster", palette="Set1")
#plot them
us_map + cluster_map


## now show the distribution of each of the clusters one at a time
## dat_pcs_mc$G is the number of clusters
par(mfrow = c(6, 2))  # Set up a 2 x 2 plotting space
output_map <- us_map
library(svMisc)
pb = txtProgressBar(min = 1, max = dat_pcs_mc$G, initial = 1)

for (g in 1:dat_pcs_mc$G) {
 setTxtProgressBar(pb,g)
 subset_dat <-subset(data,x==g)
  merged <- merge(msa_Boundary,subset_dat,by.x="NAME",by.y="X")
  #remove any cases MSAs with no cluster assignments
  merged_clean <- merged[!is.na(merged$x),]
  #convert x to factor
  merged_clean$Cluster <- as.factor(merged_clean$x)
  #convert to sf object
  cluster_sf <- st_as_sf(merged_clean)
  #make us map
  us_map <- tm_shape(us_states) + tm_borders()
  #combine them
  new_cluster_map <- tm_shape(cluster_sf)+tm_fill(col="Cluster", palette="Set1")

  #plot them
  print (us_map + new_cluster_map)
  #dev.off()#
}

```

# Examine how clusters score in different metrics
```{r, results='hide',message=FALSE}
#new data frame
dat_clust <- as.data.frame(dat)
#add names
dat_clust$ME <- rownames(dat)
rownames(dat_clust) <- NULL
#merge in cluster assignments
dat_clust <- merge(dat_clust,data,by.x="ME",by.y="X")
#calculate cluster means for different variables
c1m <- data.frame(one=colMeans(subset(dat_clust, x=="1")[2:18]))
c2m <- data.frame(two=colMeans(subset(dat_clust, x=="2")[2:18]))
c3m <- data.frame(three=colMeans(subset(dat_clust, x=="3")[2:18]))
c4m <- data.frame(four=colMeans(subset(dat_clust, x=="4")[2:18]))
c5m <- data.frame(five=colMeans(subset(dat_clust, x=="5")[2:18]))
c6m <- data.frame(six=colMeans(subset(dat_clust, x=="6")[2:18]))
c7m <- data.frame(seven=colMeans(subset(dat_clust, x=="7")[2:18]))
c8m <- data.frame(eight=colMeans(subset(dat_clust, x=="8")[2:18]))
c9m <- data.frame(nine=colMeans(subset(dat_clust, x=="9")[2:18]))
c10m <- data.frame(ten=colMeans(subset(dat_clust, x=="10")[2:18]))
c11m <- data.frame(eleven=colMeans(subset(dat_clust, x=="11")[2:18]))
cluster_means <- cbind.data.frame(c1m,c2m,c3m,c4m,c5m,c6m,c7m,c8m,c9m,c10m,c11m)
cluster_means$variable <- row.names(cluster_means)
rownames(cluster_means) <- NULL
```

```{r, fig.width=30,fig.height=15}
#boxplots of scores for clusters
par(mfrow=c(2,4), mar=c(2.5,5,2.5,3))
boxplot(subset(dat_clust, x=="1")[2:18], horizontal=T, las=1, ylim=c(-4,4), main="Cluster 1")
boxplot(subset(dat_clust, x=="2")[2:18], horizontal=T, las=1, ylim=c(-4,4), main="Cluster 2")
boxplot(subset(dat_clust, x=="3")[2:18], horizontal=T, las=1, ylim=c(-4,4), main="Cluster 3")
boxplot(subset(dat_clust, x=="4")[2:18], horizontal=T, las=1, ylim=c(-4,4), main="Cluster 4")
boxplot(subset(dat_clust, x=="5")[2:18], horizontal=T, las=1, ylim=c(-4,4), main="Cluster 5")
boxplot(subset(dat_clust, x=="6")[2:18], horizontal=T, las=1, ylim=c(-4,4), main="Cluster 6")
boxplot(subset(dat_clust, x=="7")[2:18], horizontal=T, las=1, ylim=c(-4,4), main="Cluster 7")
boxplot(subset(dat_clust, x=="8")[2:18], horizontal=T, las=1, ylim=c(-4,4), main="Cluster 8")
boxplot(subset(dat_clust, x=="9")[2:18], horizontal=T, las=1, ylim=c(-4,4), main="Cluster 9")
boxplot(subset(dat_clust, x=="10")[2:18], horizontal=T, las=1, ylim=c(-4,4), main="Cluster 10")
boxplot(subset(dat_clust, x=="11")[2:18], horizontal=T, las=1, ylim=c(-4,4), main="Cluster 11")
par(mfrow=c(1,1))
```

```{r, fig.width=30,fig.height=15}
#dotplots of cluster mean values
par(mfrow=c(2,5))
dotchart(cluster_means$one, labels=cluster_means$variable,xlim=c(-2,2), main='Cluster 1', pch=16)
abline(v=0, lwd=2)
dotchart(cluster_means$two, labels=cluster_means$variable,xlim=c(-2,2), main='Cluster 2', pch=16)
abline(v=0, lwd=2)
dotchart(cluster_means$three, labels=cluster_means$variable,xlim=c(-2,2), main='Cluster 3', pch=16)
abline(v=0, lwd=2)
dotchart(cluster_means$four, labels=cluster_means$variable,xlim=c(-2,2), main='Cluster 4', pch=16)
abline(v=0, lwd=2)
dotchart(cluster_means$five, labels=cluster_means$variable,xlim=c(-2,2), main='Cluster 5', pch=16)
abline(v=0, lwd=2)
dotchart(cluster_means$six, labels=cluster_means$variable,xlim=c(-2,2), main='Cluster 6', pch=16)
abline(v=0, lwd=2)
dotchart(cluster_means$seven, labels=cluster_means$variable,xlim=c(-2,2), main='Cluster 7', pch=16)
abline(v=0, lwd=2)
dotchart(cluster_means$eight, labels=cluster_means$variable,xlim=c(-2,2), main='Cluster 8', pch=16)
abline(v=0, lwd=2)
dotchart(cluster_means$nine, labels=cluster_means$variable,xlim=c(-2,2), main='Cluster 9', pch=16)
abline(v=0, lwd=2)
dotchart(cluster_means$ten, labels=cluster_means$variable,xlim=c(-2,2), main='Cluster 10', pch=16)
abline(v=0, lwd=2)
dotchart(cluster_means$eleven, labels=cluster_means$variable,xlim=c(-2,2), main='Cluster 11', pch=16)
abline(v=0, lwd=2)
par(mfrow=c(1,1))
```

# Which MSAs best characterize their cluster?
```{r,echo=F}
cluster1 <- data.frame(subset(dat_clust, x=="1"))
cluster2 <- data.frame(subset(dat_clust, x=="2"))
cluster3 <- data.frame(subset(dat_clust, x=="3"))
cluster4 <- data.frame(subset(dat_clust, x=="4"))
cluster5 <- data.frame(subset(dat_clust, x=="5"))
cluster6 <- data.frame(subset(dat_clust, x=="6"))
cluster7 <- data.frame(subset(dat_clust, x=="7"))
cluster8 <- data.frame(subset(dat_clust, x=="8"))
cluster9 <- data.frame(subset(dat_clust, x=="9"))
cluster10 <- data.frame(subset(dat_clust, x=="10"))
cluster11 <- data.frame(subset(dat_clust, x=="11"))

#function to compute smallest combined absolute difference from mean
Average_MSA_inClust <- function(x) {
  output = data.frame(Name=x$ME)
  for (i in 2:ncol(x)) {
    output[[i]] <- abs(x[[i]] - mean(x[[i]]))
  }
  output$s_diff = rowSums(output[2:ncol(output)])
  cluster_average = output[which.min(output$s_diff),]
  return(cluster_average)
}

cluster1_average <- Average_MSA_inClust(cluster1)
cluster2_average <- Average_MSA_inClust(cluster2)
cluster3_average <- Average_MSA_inClust(cluster3)
cluster4_average <- Average_MSA_inClust(cluster4)
cluster5_average <- Average_MSA_inClust(cluster5)
cluster6_average <- Average_MSA_inClust(cluster6)
cluster7_average <- Average_MSA_inClust(cluster7)
cluster8_average <- Average_MSA_inClust(cluster8)
cluster9_average <- Average_MSA_inClust(cluster9)
cluster10_average <- Average_MSA_inClust(cluster10)
cluster11_average <- Average_MSA_inClust(cluster11)
typical_members <- data.frame(Cluster=c(1:11), Name=c(cluster1_average$Name,
                              cluster2_average$Name,
                              cluster3_average$Name,
                              cluster4_average$Name,
                              cluster5_average$Name,
                              cluster6_average$Name,
                              cluster7_average$Name,
                              cluster8_average$Name,
                              cluster9_average$Name,
                              cluster10_average$Name,
                              cluster11_average$Name))

kable(typical_members, caption = "Typical members of each cluster")
```

# Which cluster scores the best on the different metrics?
```{r, echo=F}
##########################
#Most sustainable cluster

#rotate columns and rows
cluster_means_t <- cluster_means[1:11]
row.names(cluster_means_t) <- cluster_means$variable
cluster_means_t <- data.frame(t(cluster_means_t))
cluster_means_t$cluster <- row.names(cluster_means_t)
row.names(cluster_means_t) <- NULL

#clusters scoring the best in the different metrics
best_air <- cluster_means_t[which.max(cluster_means_t$AQI_Good),]
best_edu <- cluster_means_t[which.max(cluster_means_t$Bachelor_Over_25),]
best_pov <- cluster_means_t[which.min(cluster_means_t$Per_Poverty),]
best_ineq <- cluster_means_t[which.min(cluster_means_t$Gini),]
best_hous <- cluster_means_t[which.min(cluster_means_t$Per_Sev_Hous),]
best_stream <- cluster_means_t[which.min(cluster_means_t$Xstreamlengthimpaired),]
best_land <- cluster_means_t[which.max(cluster_means_t$Per_Avg_Land_Cov),]
best_health <- cluster_means_t[which.min(cluster_means_t$poor_health_percent),]
best_water <- cluster_means_t[which.min(cluster_means_t$Z_Water_Index),]
low_index_black <- cluster_means_t[which.min(cluster_means_t$Black_equality),]
#low_index_asian <- cluster_means_t[which.min(cluster_means_t$Asian_equality),]
low_index_latino <- cluster_means_t[which.min(cluster_means_t$Hisspanic_equality),]
best_emissions <- cluster_means_t[which.min(cluster_means_t$GHG_Percap),]
best_employmet <- cluster_means_t[which.min(cluster_means_t$UNEMPLOY),]
best_food <- cluster_means_t[which.min(cluster_means_t$FOOD_INS),]
best_crime <- cluster_means_t[which.min(cluster_means_t$VIO_CRIME),]
best_migration <-cluster_means_t[which.min(cluster_means_t$non_migration),]
best_broadband <-cluster_means_t[which.min(cluster_means_t$broadband_access),]
#best_female <-cluster_means_t[which.min(cluster_means_t$Female_Inequality),]

most_sustainable_clusters <- data.frame(Metric=c(colnames(cluster_means_t[c(1:16)])),
                                        Cluster=c(
                                        best_air$cluster,
                                        best_edu$cluster,
                                        best_pov$cluster,
                                        best_ineq$cluster,
                                        best_hous$cluster,
                                        best_stream$cluster,
                                        best_land$cluster,
                                        best_health$cluster,
                                        best_water$cluster,
                                        low_index_black$cluster,
                                        #low_index_asian$cluster,
                                        low_index_latino$cluster,
                                        best_emissions$cluster,
                                        best_employmet$cluster,
                                        best_food$cluster,
                                        best_crime$cluster,
                                        #best_female$cluster,
                                        best_broadband$cluster,
                                        best_migration$cluster))

kable(most_sustainable_clusters)
```

## Table showing which clusters score best and worst for different metrics
```{r, }
cluster_means_t %>%
  kbl() %>%
  kable_paper(full_width=T) %>%
  #column_spec(1, color = ifelse(cluster_means_t$AQI_Good < max(cluster_means_t$AQI_Good), 'black','blue'))
  column_spec(1, color = ifelse(cluster_means_t[,1] == max(cluster_means_t[,1]), "blue",
                                   ifelse(cluster_means_t[,1] == min(cluster_means_t[,1]), "red", "black"))) %>%
  column_spec(2, color = ifelse(cluster_means_t[,2] == max(cluster_means_t[,2]), "blue",
                                   ifelse(cluster_means_t[,2] == min(cluster_means_t[,2]), "red", "black"))) %>%
  column_spec(8, color = ifelse(cluster_means_t[,8] == max(cluster_means_t[,8]), "blue",
                                   ifelse(cluster_means_t[,8] == min(cluster_means_t[,8]), "red", "black"))) %>%
  column_spec(3, color = ifelse(cluster_means_t[,3] == max(cluster_means_t[,3]), "red",
                                   ifelse(cluster_means_t[,3] == min(cluster_means_t[,3]), "blue", "black"))) %>%
  column_spec(4, color = ifelse(cluster_means_t[,4] == max(cluster_means_t[,4]), "red",
                                   ifelse(cluster_means_t[,4] == min(cluster_means_t[,4]), "blue", "black"))) %>%
  column_spec(5, color = ifelse(cluster_means_t[,5] == max(cluster_means_t[,5]), "red",
                                   ifelse(cluster_means_t[,5] == min(cluster_means_t[,5]), "blue", "black"))) %>%
  column_spec(6, color = ifelse(cluster_means_t[,6] == max(cluster_means_t[,6]), "red",
                                   ifelse(cluster_means_t[,6] == min(cluster_means_t[,6]), "blue", "black"))) %>%
  column_spec(7, color = ifelse(cluster_means_t[,7] == max(cluster_means_t[,7]), "red",
                                   ifelse(cluster_means_t[,7] == min(cluster_means_t[,7]), "blue", "black"))) %>%
  column_spec(9, color = ifelse(cluster_means_t[,9] == max(cluster_means_t[,9]), "red",
                                   ifelse(cluster_means_t[,9] == min(cluster_means_t[,9]), "blue", "black"))) %>%
  column_spec(10, color = ifelse(cluster_means_t[,10] == max(cluster_means_t[,10]), "red",
                                   ifelse(cluster_means_t[,10] == min(cluster_means_t[,10]), "blue", "black"))) %>%
  column_spec(11, color = ifelse(cluster_means_t[,11] == max(cluster_means_t[,11]), "red",
                                   ifelse(cluster_means_t[,11] == min(cluster_means_t[,11]), "blue", "black"))) %>%
  column_spec(12, color = ifelse(cluster_means_t[,12] == max(cluster_means_t[,12]), "red",
                                   ifelse(cluster_means_t[,12] == min(cluster_means_t[,12]), "blue", "black"))) %>%
  column_spec(13, color = ifelse(cluster_means_t[,13] == max(cluster_means_t[,13]), "red",
                                   ifelse(cluster_means_t[,13] == min(cluster_means_t[,13]), "blue", "black"))) %>%
  column_spec(14, color = ifelse(cluster_means_t[,14] == max(cluster_means_t[,14]), "red",
                                   ifelse(cluster_means_t[,14] == min(cluster_means_t[,14]), "blue", "black"))) %>%
  column_spec(15, color = ifelse(cluster_means_t[,15] == max(cluster_means_t[,15]), "red",
                                   ifelse(cluster_means_t[,15] == min(cluster_means_t[,15]), "blue", "black"))) %>%
  column_spec(16, color = ifelse(cluster_means_t[,16] == max(cluster_means_t[,16]), "red",
                                   ifelse(cluster_means_t[,16] == min(cluster_means_t[,16]), "blue", "black"))) %>%
  column_spec(17, color = ifelse(cluster_means_t[,17] == max(cluster_means_t[,17]), "red",
                                   ifelse(cluster_means_t[,17] == min(cluster_means_t[,17]), "blue", "black")))
  
```


